---
title: "Mapping urban-rural balance sentiment in Lexington/Fayette County, KY"
author: 
  - name: "Conley Kinkead"
    url: "https://conleykinkead.github.io/"
    affiliations:
      - name: "University of Kentucky"
knitr: 
  opts_chunk: 
    message: false
    warning: false
    fig-width: 5
    fig-height: 3
    fig-align: center
    cap-location: margin
execute:
  freeze: auto  # re-render only when source changes
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: body
    toc-title: "Contents"
    theme: 
      light: [flatly, custom.scss]
      dark: [darkly, custom.scss]
    title-block-banner: true
    include-in-header: google-font.html
    mainfont: Quicksand
  pdf: default
pdf-engine: pdflatex
---

## Project introduction

This project analyzes and maps urban-rural balance sentiment in Lexington/Fayette County, Kentucky using data from CivicLex's [On The Table (OTT)](https://www.ottlex.org/) survey data, which was conducted in the spring of 2022.

Survey participants completed both open response and multiple choice questions covering a variety of topics relevant to Lexington. The urban-rural balance topic is significant to Lexington as the city continues to evaluate making changes to its [Urban Service Area](https://conleykinkead.github.io/lexington_urban-service-area/index.html) and is economically and culturally intertwined with the thoroughbred horse industry, but is also growing in population and other industry within the city's urban core. A sentiment analysis of urban-rural balance survey response data can shed light on how residents feel, ranging from very negative to very positive, about Lexington/Fayette County's management of and balance between urban and rural land here. In this project, sentiment on this topic is measured and then mapped based on where respondents live. I also delve into the content of responses to look at important words or phrases in the open responses.

Importantly, participation in OTT was entirely voluntary, so the data and the following results shouldn't be assumed to be representative of Lexington/Fayette County residents at large. CivicLex determined the city council districts that respondents are likely to reside in based on their self-identified neighborhood of residence; thus the following analysis shows "likely council district" instead of "council district".

All data and files associated with this analysis can be found on the [project Github repository](https://github.com/conleykinkead/ottlex_urbalancesentiment).

## Prepping for analysis

### Installing packages

```{r}
#| eval: false

install.packages("tidyverse")
install.packages("tidytext")
install.packages("janitor")
install.packages("fs")
install.packages("here")
install.packages("sf")
install.packages("tmap")
install.packages("RColorBrewer")
install.packages("quanteda")
install.packages("quanteda.textplots")
install.packages("quanteda.textstats")
install.packages("textdata")
```

### Loading packages

```{r}
library(tidyverse)
library(janitor)
library(fs)
library(here)
library(tidytext)
library(sf)
library(tmap)
library(RColorBrewer)
library(quanteda)
library(quanteda.textplots)
library(quanteda.textstats)
library(textdata)
```

## OTT data

### Downloading CivicLex On The Table (OTT) data

```{r}
#| eval: false

dir_create(here("data", "zipfiles"))
download.file(
  "https://www.ottlex.org/s/OTT-2022-Raw-Data.zip",
  destfile = here("data", "zipfiles", "ott.zip")
  )
```

```{r}
#| eval: false

dir_create(here("data", "ott")) 
unzip(
  zipfile = here("data", "zipfiles", "ott.zip"),
  junkpaths = T,
  exdir = here("data", "ott")
)
```

### Tidying OTT data

```{r}
ott_data <- read_csv(here("data", "ott", "OTT Raw Data-Raw Data Grid.csv")) |> 
  clean_names()
```

### Selecting the required data

```{r}
ott_data |>
  select(
    ott_raw_data_response,
    likely_council_district,
    ur_balance_or
    )|>
  pivot_longer(
    ur_balance_or,
    values_to = "response"  
) |>
  mutate(
    response = str_squish(response)
  ) |>
  filter(
    str_length(response) > 0
  ) |>
  mutate(
    docid = str_c(ott_raw_data_response, sep = "_")
  ) -> ur_balance_tidy
```

## Sentiment analysis of Lexington's urban-rural balance

### Combining urban-rural balance open responses by likely council district

```{r}
ur_balance_tidy |>
    filter(
    str_length(likely_council_district) < 3
    ) |>
  group_by(likely_council_district) |>
  summarise(text = str_flatten(response, collapse = " ")) -> district_response
```

### Measuring urban-rural balance sentiment by likely council district

#### Tokenizing responses

```{r}
district_response |>
  unnest_tokens(
    word,
    text,
    token = "words"
  ) -> district_response_tokens
```

#### Joining tokens with AFFIN lexicon

[AFFIN](https://search.r-project.org/CRAN/refmans/textdata/html/lexicon_afinn.html) is a dataset containing a lexicon of English words rated -5 to +5 based on their sentiment, with -5 being the most negative, 0 being neutral, and +5 being the most positive.

```{r}
district_response_tokens |>
  mutate(
    word_location = row_number()
  ) |>
  inner_join(
    get_sentiments("afinn")
  ) -> ur_sentiments
```

#### Grouping and averaging sentiment by likely council district

```{r}
ur_sentiments |>
  group_by(likely_council_district) |>
  summarise(
    sentiment = mean(value)
  ) -> ur_district_sentiment
```

#### Plot of urban-rural balance sentiment by likely council district

```{r}
#| fig-cap: "Rough urban-rural balance sentiment for likely council districts plot to begin visualizing the data."
#| fig-width: 4
#| fig-height: 3

ur_district_sentiment |>
  ggplot(aes(likely_council_district, sentiment)) +
  geom_point()
```

## Mapping sentiment by likely council district

### Adding council district spatial data

#### Downloading [council district GeoJSON](https://data.lexingtonky.gov/datasets/5bab826b97c84d229f7abf0bc947981c_0/explore?location=38.027862%2C-84.471932%2C11.56) from [Lexington's Data Hub](https://data.lexingtonky.gov/)

```{r}
lex_councils <- read_sf("https://services1.arcgis.com/Mg7DLdfYcSWIaDnu/arcgis/rest/services/Council_District/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")
```

### Joining council district spatial df with sentiment analysis df

```{r}
ur_district_sentiment |>
  mutate(
    DISTRICT = as.numeric(likely_council_district)
  ) -> ur_district_sentiment_dbl
```

```{r}
district_sentiment <-
  merge(ur_district_sentiment_dbl,lex_councils)
```

#### Making merged df a spatial df

```{r}
district_sentiment_sf = st_as_sf(district_sentiment)
```

## Map of urban-rural balance sentiment

```{r}
#| fig-cap: "Urban-rural balance sentiment by likely council district map. A higher sentiment value represents a more positive sentiment."
#| fig-width: 8
#| fig-height: 8

tm_shape(district_sentiment_sf) +
  tm_polygons(col = "sentiment", palette = "Blues") +
  tm_credits("Data sources: CivicLex; LFUCG", size = 0.9, position = c("left","top")) +
  tm_credits("Author: Conley Kinkead", size = 0.9, position = c("left","top")) +
  tm_scale_bar(position = c("right", "bottom")) +
  tm_layout(main.title = "Urban-rural balance sentiment by likely council district",
            fontface = "italic",
            fontfamily = "serif",
            main.title.size = 1.5) +
  tm_legend(
  legend.title.size = 1.2,
  legend.title.fontface = "bold",
  legend.text.size = 0.9,
  legend.bg.color = "white",
  legend.position = c("left", "bottom"))
```

### Results

This map shows that residents of all council districts have a positive sentiment regarding Lexington's urban-rural balance, but at varying degrees. AFINN ratings range from -5 (very negative) to +5 (very positive), with 0 indicating a neutral sentiment. A lower sentiment score on this map indicates a less positive sentiment, and a higher score indicates a more positive sentiment.

The results showing that the OTT participants who are likely residents of council district 12 have the least positive sentiment regarding Lexington's urban-rural balance is interesting. District 12 covers the primarily agricultural area of Lexington/Fayette County, wrapping around the city's central area, and largely overlapping with the Rural Service Area.

## A little more analysis: Urban-rural balance response content analysis using quanteda

```{r}
ur_balance_tidy |>
  corpus(
    docid_field = "docid",
    text_field = "response"
  ) -> ur_balance_corpus
```

#### Corpus to tokens

```{r}
ur_balance_corpus |>
  tokens(remove_punct = T) |>
  tokens_tolower() |>
  tokens_remove(pattern = stopwords()) |>
  tokens_compound(pattern = list(c("man", "o", "war"), c("horse", "capital"))) -> ur_balance_tokens
```

#### Tokens to DFM

```{r}
ur_balance_tokens |>
  dfm() -> ur_balance_dfm
```

### Keyness plot of likely council district 12 urban-rural balance response content versus all others

```{r}
#| fig-cap: "Keyness plot of urban-rural balance response content. This plot shows the words that are key for differentiating urban-rural balance open response content for likely residents of council district 12 versus all other council districts in Lexington/Fayette County."
#| fig-width: 8
#| fig-height: 8

ur_balance_dfm |>
  dfm_group(groups = ur_balance_dfm$likely_council_district) |>
  textstat_keyness(target = "12") |>
  textplot_keyness()
```

### Results

This keyness plot shows words or phrases that respondents who are likely residents of council district 12 used more often in comparison to respondents who are likely residents of all other council districts combined. The words and phrases plotted in blue are the key differentiating terms for likely council district 12 urban-rural balance open responses, while the words plotted in grey represent the key differential words for the same responses for likely residents of all other council districts combined.
